### 2017-08-30 reports/thesis/book.tex
In thesis, added M step.

The EM algorithm \citep{dempster1977maximum} instead treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{Y_1,Y_2,Y_3,\dotdotdot, Y_n\right\}$ be realizations of $Y$ and $\left\{x_1,x_2,x_3,\dotdotdot, x_n\right\}$ be measurements of $X|Y$. Define the joint log likelihood to be
\begin{multline*}
\ln L(\lambda,\alpha,\beta;X,Y)
=
\sum_{i=1}^n
\left[
	\mathbb{I}(Y_i=0)
	\ln
	\prob(Y=0)
	\right.\\\left.
	+
	\mathbb{I}(Y_i>0)
	\ln
	\left[
	p_{X|Y}(x_i|Y_i)\prob(Y=Y_i)
	\right]
\right]
\end{multline*}
so that
\begin{multline}
\ln L(\lambda,\alpha,\beta;X,Y)
=
\sum_{i=1}^n
\left[
	-\mathbb{I}(Y_i=0)
	\lambda
	\right.\\\left.+
	\mathbb{I}(Y_i>0)
	\left(
	Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
	\right.\right.\\\left.\left.	
	- \lambda + Y_i \ln \lambda - \ln(Y!)
	\right)
\right] \ .
\end{multline}
The estimators are found by optimising the joint log likelihood. This is done iteratively by estimating the $Y$s given the $X$s and parameters (E step), followed by estimating the parameters given the $Y$s and $X$s (M step) until some convergence conditions are met.

@@

Let
\begin{equation}
W_y^{(r)} = y^r W_y
\end{equation}
so that
\begin{equation}
y_i = \frac{\sum_{y=1}^\infty W^{(1)}_y}{\sum_{y=1}^\infty W_y} \ .
\end{equation}
The exponent $r=0,1,2,\dotdotdot$ will be useful in the M step where higher conditional moments are needed.
The sum can be truncated
\begin{equation}
\sum_{y=1}^\infty W^{(r)}_y \approx \sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y
\end{equation}
where $y_\text{l}<y_\text{max}<y_\text{u}$ and $y_\text{max}$ is the value of $y$ which maximises $W_y^{(r)}$. By expressing $\ln(W^{(r)}_y)$ as
\begin{equation}
\ln\left(W_y^{(r)}\right)=r\ln(y)+\ln(W_y)
\end{equation}
and then taking the derivative with respect to $y$
\begin{equation}
\frac{\partial \ln(W_y^{(r)})}{\partial y} = \frac{r}{y } + \frac{\partial \ln(W_y)}{\partial y} \ .
\end{equation}
Keep in mind that $y=1,2,3,\dotdotdot$ so for large $y$, an approximation can be made $r/y\approx 0$ so that
\begin{equation}
\frac{\partial \ln(W_y^{(r)})}{\partial y} \approx \frac{\partial \ln(W_y)}{\partial y} \ .
\end{equation}
By using such an approximation, the method for evaluating the sum $\sum_{y=1}^\infty W^{(r)}_y$ is almost exactly the same as evaluating the sum $\sum_{y=1}^\infty W_y$, in particular $y_\text{max}$ are the same. For a given $r=0,1,2,\dotdotdot$, the limit of the sum $\sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y$ are chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}^{(r)}$ where $\epsilon$ is some small constant. The limits will be different for different values of $r$.

@@

>>>>

The EM algorithm \citep{dempster1977maximum} instead treats $Y\sim\poisson(\lambda)$ as a latent variable. Let $\left\{Y_1,Y_2,Y_3,\dotdotdot, Y_n\right\}$ be realizations of $Y$ and $\left\{x_1,x_2,x_3,\dotdotdot, x_n\right\}$ be measurements of $X|Y$. Define the joint log likelihood to be
\begin{multline*}
\ln L(\lambda,\alpha,\beta;X,Y)
=
\sum_{i=1}^n
\left[
	\mathbb{I}(x_i=0)
	\ln
	\prob(Y=0)
	\right.\\\left.
	+
	\mathbb{I}(x_i>0)
	\ln
	\left[
	p_{X|Y}(x_i|Y_i)\prob(Y=Y_i)
	\right]
\right]
\end{multline*}
so that
\begin{multline}
\ln L(\lambda,\alpha,\beta;X,Y)
=
\sum_{i=1}^n
\left[
	-\mathbb{I}(x_i=0)
	\lambda
	\right.\\\left.+
	\mathbb{I}(x_i>0)
	\left(
	Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
	\right.\right.\\\left.\left.	
	- \lambda + Y_i \ln \lambda - \ln(Y_i!)
	\right)
\right] \ .
\end{multline}
The estimators are found by optimising the joint log likelihood. This is done iteratively by estimating the $Y$s given the $X$s and parameters (E step), followed by estimating the parameters given the $Y$s and $X$s (M step) until some convergence conditions are met.

@@

Let
\begin{equation}
W_y^{(r)} = y^r W_y
\end{equation}
so that
\begin{equation}
y_i = \frac{\sum_{y=1}^\infty W^{(1)}_y}{\sum_{y=1}^\infty W_y}
\end{equation}
and
\begin{equation}
\variance[Y|X=x_i] = \frac{\sum_{y=1}^\infty W^{(2)}_y}{\sum_{y=1}^\infty W_y} - \left(y_i\right)^2 \ .
\end{equation}
The exponent $r=0,1,2,\dotdotdot$ will be useful in the M step where higher conditional moments are needed.
The sum can be truncated
\begin{equation}
\sum_{y=1}^\infty W^{(r)}_y \approx \sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y
\end{equation}
where $y_\text{l}<y_\text{max}<y_\text{u}$ and $y_\text{max}$ is the value of $y$ which maximises $W_y^{(r)}$. By expressing $\ln(W^{(r)}_y)$ as
\begin{equation}
\ln\left(W_y^{(r)}\right)=r\ln(y)+\ln(W_y)
\end{equation}
and then taking the derivative with respect to $y$
\begin{equation}
\frac{\partial \ln(W_y^{(r)})}{\partial y} = \frac{r}{y } + \frac{\partial \ln(W_y)}{\partial y} \ .
\end{equation}
Keep in mind that $y=1,2,3,\dotdotdot$ so for large $y$, an approximation can be made $r/y\approx 0$ so that
\begin{equation}
\frac{\partial \ln(W_y^{(r)})}{\partial y} \approx \frac{\partial \ln(W_y)}{\partial y} \ .
\end{equation}
By using such an approximation, the method for evaluating the sum $\sum_{y=1}^\infty W^{(r)}_y$ is almost exactly the same as evaluating the sum $\sum_{y=1}^\infty W_y$, in particular $y_\text{max}$ are the same. For a given $r=0,1,2,\dotdotdot$, the limit of the sum $\sum_{y=y_\text{l}}^{y_\text{u}} W^{(r)}_y$ are chosen such that $W_{y_\text{l}}$ and $W_{y_\text{u}}$ are less than $\epsilon W_{y_\text{max}}^{(r)}$ where $\epsilon$ is some small constant. The limits will be different for different values of $r$.

@@

In the M step, the conditional expected joint log likelihood is maximised with respect to the parameters $\lambda$, $\alpha$ and $\beta$. The objective function is then
\begin{multline*}
T(\lambda,\alpha,\beta)
=
\sum_{i=1}^n
\expectation\left[
	-\mathbb{I}(x_i=0)
	\lambda
	\right.\\\left.+
	\mathbb{I}(x_i>0)
	\left(
	Y_i\alpha\ln\beta-\ln\Gamma(Y_i\alpha)+(Y_i\alpha-1)\ln x_i - \beta x_i
	\right.\right.\\\left.\left.	
	- \lambda + Y_i \ln \lambda - \ln(Y_i!)
	\right)
|X_i=x_i\right]
\end{multline*}
\begin{multline*}
T(\lambda,\alpha,\beta)
=
-n\lambda
\\+
	\sum_{i=1}^n
	\mathbb{I}(x_i>0)
	\left[
	\expectation[Y_i|X_i=x_i]\alpha\ln\beta-\expectation[\ln\Gamma(Y_i\alpha)|X_i=x_i]
	\right.\\\left.
	+\expectation[Y_i|X_i=x_i]\alpha\ln x_i - \beta x_i
	+ \expectation[Y_i|X_i=x_i] \ln \lambda
	\right] + c
\end{multline*}
where $c$ is some constant not dependent on $\lambda$, $\alpha$ or $\beta$.

The conditional expectation $y_i = \expectation[Y_i|X_i=x_i]$ was calculated in the E step. The quantity $\expectation[\ln\Gamma(Y_i\alpha)|X_i=x_i]$ can be calculated using the approximation
\begin{equation}
\expectation[\ln\Gamma(Y_i\alpha)|X_i=x_i] \approx
\ln\Gamma(\alpha y_i) + \frac{1}{2}\zeta_i\alpha^2\psi'(y_i\alpha)
\end{equation}
where
\begin{equation}
\zeta_i = \variance[Y_i|X_i=x_i]
\end{equation}
calculated in the E step and $\psi$ is the digamma function. The objective function is then
\begin{multline}
T(\lambda,\alpha,\beta)
=
-n\lambda+
	\sum_{i=1}^n
	\mathbb{I}(x_i>0)
	\left[
	y_i\alpha\ln\beta-\ln\Gamma(\alpha y_i) - \frac{1}{2}\zeta_i\alpha^2\psi'(y_i\alpha)
	\right.\\\left.
	+y_i\alpha\ln x_i - \beta x_i
	+ y_i \ln \lambda
	\vphantom{\frac{1}{1}}\right] + c \ .
\end{multline}

Taking the derivative with respect to $\lambda$
\begin{equation}
\frac{\partial T}{\partial\lambda} = -n + \frac{\sum_{i=1}^ny_i}{\lambda}
\end{equation}
and setting it to zero
\begin{equation}
\widehat{\lambda} = \frac{\sum_{i=1}^n y_i}{n}
\end{equation}
obtains a M step estimator, and a familiar one, for $\lambda$. Taking the second derivative with respect to $\lambda$
\begin{equation}
\frac{\partial^2 T}{\partial \lambda^2} = -\frac{\sum_{i=1}^ny_i}{\lambda^2} < 0
\end{equation}
verifies $\widehat{\lambda}$ maximises T. In addition
\begin{equation}
\frac{\partial^2 T}{\partial \alpha \partial \lambda } = 0
\end{equation}
and
\begin{equation}
\frac{\partial^2 T}{\partial \beta \partial \lambda } = 0 \ .
\end{equation}

Maximising $T$ with respect to $\alpha$ and $\beta$ can be done numerically using the Newton-Raphson method since derivatives up to the second order can be obtained. For the first order these are:
\begin{multline}
\frac{\partial T}{\partial \alpha} = \sum_{i=1}^n\mathbb{I}(x_i>0)\left[
	y_i\ln\beta -\psi(\alpha y_i)y_i-\zeta_i\alpha\psi'(\alpha y_i)
	\vphantom{\frac{1}{1}}\right.\\\left.
	- \frac{1}{2}\zeta_i\alpha^2\psi''(\alpha y_i)y_i + y_i\ln x_i
\right]
\end{multline}
and
\begin{equation}
\frac{\partial T}{\partial \beta} = \sum_{i=1}^n\mathbb{I}(x_i>0)\left[
\frac{\alpha y_i}{\beta}-x_i
\right] \ .
\end{equation}
The second orders are:
\begin{equation}
\frac{\partial^2 T}{\partial \alpha \partial \beta} =
\sum_{i=1}^n \mathbb{I}(x_i>0)\left[\frac{y_i}{\beta}\right] \ ,
\end{equation}
\begin{equation}
\frac{\partial^2 T}{\partial \beta^2} = \sum_{i=1}^n\mathbb{I}(x_i>0)\left[-\frac{\alpha y_i}{\beta^2}
\right]
\end{equation}
and
\begin{multline*}
\frac{\partial^2 T}{\partial \alpha^2} =  \sum_{i=1}^n\mathbb{I}(x_i>0)\left[
	-y_i^2\psi'(\alpha y_i) - \zeta_i\psi'(\alpha y_i) - \zeta_i\alpha y_i\psi''(\alpha y_i)
	\vphantom{\frac{1}{2}}\right.\\\left.	
	- \zeta_i\alpha\psi''(\alpha y_i)y_i
	-\frac{1}{2}\zeta_i\alpha^2\psi'''(\alpha y_i)y_i^2
\right]
\end{multline*}
simplifying to
\begin{multline}
\frac{\partial^2 T}{\partial \alpha^2} =  \sum_{i=1}^n\mathbb{I}(x_i>0)\left[
	-(y_i^2+\zeta_i)\psi'(\alpha y_i) - 2\zeta_i\alpha y_i\psi''(\alpha y_i)
	\vphantom{\frac{1}{2}}\right.\\\left.	
	-\frac{1}{2}\zeta_i\alpha^2y_i^2\psi'''(\alpha y_i)
\right] \ .
\end{multline}

All the derivatives can be used in the Newton-Raphson iterative update, which is
\begin{equation}
\begin{pmatrix}
\alpha \\ \beta
\end{pmatrix}
\leftarrow
\begin{pmatrix}
\alpha \\ \beta
\end{pmatrix}
\left[
	\nabla_{\alpha,\beta}\nabla_{\alpha,\beta}\T T
\right]^{-1}
\left[\nabla_{\alpha,\beta} T\right]
\end{equation}
where
\begin{equation}
\nabla_{\alpha,\beta}=
\begin{pmatrix}
	{\partial}/{\partial \alpha}
	\\
	{\partial}/{\partial \beta}
\end{pmatrix} \ .
\end{equation}
Since increasing $T$ is sufficient for the EM algorithm \citep{dempster1977maximum}, one step of the Newton-Raphson iterative update was chosen in the M step to avoid implementing a convergence condition.
